What words ought to exist? \\ {\normalsize Coining with coinduction}

### 2011-03-20 trunk/wishlist/paper.tex
first attempt at backformation

\vspace{0.2in}
\begin{tabular}{rl@{\quad\quad}rl}
22.2\%  &  hel      &      2.5\%  &  helpfel  \\
11.1\%  &  helo     &      2.5\%  &  helafel  \\
 7.4\%  &  fel      &      1.9\%  &  fulo     \\
 3.7\%  &  hell     &      1.9\%  &  hello    \\
 3.7\%  &  felo     &      1.2\%  &  fell     \\
 3.7\%  &  ful      &      1.2\%  &  helafelo \\
\end{tabular}
\vspace{0.2in}

@@

The next step is to build a Markov chain for a list of real words and
see what results. I built one for the SOWPODS word list, and these
are the most likely words, with real words filtered out:

% XXX improve layout
\vspace{0.2in}
\begin{tabular}{rl@{\quad\quad}rl}
0.049901545  & s     &   0.001768358  & y   \\
0.017526947  & d     &   0.001761301  & p   \\
0.009521421  & g     &   0.001653041  & a   \\
0.005506083  & c     &   0.001652864  & n   \\
0.004376836  & r     &   0.001563810  & ps  \\
0.004218223  & t     &   0.001383503  & ms  \\
0.004088377  & e     &   0.001371545  & ts  \\
0.003560976  & m     &   0.001327023  & ds  \\
0.003248559  & ss    &   0.001188715  & hy  \\
0.002001529  & rs    &   0.001166457  & k   \\
0.001958674  & h     &   0.001160619  & ng  \\
0.001873761  & l     &   0.001142785  & ly  \\
\end{tabular}
\vspace{0.2in}

@@

There is a standard solution to this problem, which is to generalize
the Markov chain to keep more than one letter of history. So instead
of just tallying how often {\it g} follows {\it n}, we count how often
{\it g} follows {\it in} (and any other pair of letters). This makes
the table pretty large, but I went through some trouble to put a
graphic in Figure~\ref{fig:sowpodstrigrams} because it looks pretty
cool.

@@

most people remember induction, perhaps

coinduction: never gonna give you up. once you pop you can't stop.

@@

% \section{Ockham's Regexp?}

@@

>>>>

\begin{center}
\begin{tabular}{rl@{\quad\quad}rl}
22.2\%  &  hel      &      2.5\%  &  helpfel  \\
11.1\%  &  helo     &      2.5\%  &  helafel  \\
 7.4\%  &  fel      &      1.9\%  &  fulo     \\
 3.7\%  &  hell     &      1.9\%  &  hello    \\
 3.7\%  &  felo     &      1.2\%  &  fell     \\
 3.7\%  &  ful      &      1.2\%  &  helafelo \\
\end{tabular}
\end{center}

@@

\begin{figure}
\includegraphics[width=\linewidth]{coin1}
\caption{Markov chain for the SOWPODS word list, where darker squares
indicate higher probability. The darkest is the transition from {\it q}
to {\it u} ($98\%$), which is not surprising.}
\label{fig:sowpodsbigrams}
\end{figure}

The next step is to build a Markov chain for a list of real words and
see what results. I built one for the SOWPODS word list, which results
in the table in Figure~\ref{fig:sowpodsbigrams}. These
are the most likely words, with real words filtered out:

% XXX improve layout
\begin{center}
\begin{tabular}{rl@{\quad\quad}rl}
0.049901545  & s     &   0.001768358  & y   \\
0.017526947  & d     &   0.001761301  & p   \\
0.009521421  & g     &   0.001653041  & a   \\
0.005506083  & c     &   0.001652864  & n   \\
0.004376836  & r     &   0.001563810  & ps  \\
0.004218223  & t     &   0.001383503  & ms  \\
0.004088377  & e     &   0.001371545  & ts  \\
0.003560976  & m     &   0.001327023  & ds  \\
0.003248559  & ss    &   0.001188715  & hy  \\
0.002001529  & rs    &   0.001166457  & k   \\
0.001958674  & h     &   0.001160619  & ng  \\
0.001873761  & l     &   0.001142785  & ly  \\
\end{tabular}
\end{center}

@@

There is a standard solution to this problem, which is to generalize
the Markov chain to keep more than one letter of history. So instead
of just tallying how often {\it g} follows {\it n}, we count how often
{\it g} follows {\it in} (and any other pair of letters).\!\footnote{
  The details are straightforward, except possibly that we now imagine
  each word to start with two (or in general, $n$) copies of the start
  symbol, so that we see ``\<\<helpful\>''. The column corresponding
  to the history {\bf \<\<} tells us the frequency of letters that
  start words, and for example the column {\bf \<h} tells us the
  frequency of letters that follow {\it h} when it appears at the
  start of a word. We do not need to repeat the ending character \>
  because once we see it, we never do anything but end the word.} This
makes the table pretty large, so you'll just have to look at
Figure~\ref{fig:sowpodsbigrams} again and imagine it being 28 times
wider. But the good news is that it invents much better words:

\begin{verbatim}
Markov chain with n=2.
100 top paths:
0.007092576 ing
0.002483309 ses
0.001694275 des
0.001548279 nes
0.001403047 sts
0.001314279 se
0.001287353 ings
0.001262904 ded
0.001170356 cal
0.001101124 le
0.001078157 der
0.001070769 ove
0.001016514 gly
0.000882527 hy
0.000853499 ung
0.000838535 cy
0.000816757 pres
0.000802384 pers
0.000787731 ps
0.000771953 ders
0.000746855 ly
0.000732157 ines
0.000709460 ve
0.000650273 sm
0.000650062 ty
0.000642246 pred
0.000627432 ble
0.000622587 hed
0.000603012 ter
0.000587263 gy
0.000564629 sms
0.000549088 py
0.000545497 ans
0.000541832 ber
0.000531745 ce
0.000489139 ang
0.000487316 unt
0.000482908 snes
\end{verbatim}

\begin{verbatim}
Markov chain with n=3.
100 top paths:
0.001093377 des
0.000789743 pers
0.000767821 cal
0.000622381 pres
0.000452415 nons
0.000449855 ress
0.000426892 ing
0.000404454 pred
0.000387570 ent
0.000366753 dist
0.000357115 ble
0.000353576 ches
0.000348642 gly
0.000345564 inted
0.000340691 dists
0.000339750 lity
0.000336468 und
0.000331914 dises
0.000331389 int
0.000320104 ints
0.000291846 ching
0.000284350 tely
0.000284228 dise
0.000279669 ster
0.000279431 ves
0.000276676 stic
0.000272655 tric
0.000263922 cally
0.000261415 ter
0.000260898 ents
0.000258270 rous
0.000253990 dising
0.000245834 phy
0.000244272 dised
0.000240505 hous
0.000239726 resses
0.000237100 sters
0.000236753 ove
0.000230976 dism
0.000227701 ched
0.000225104 fing
0.000222993 ence
0.000221624 ters
0.000219511 disms
0.000219246 unders
\end{verbatim}

\begin{verbatim}
Markov chain with n=4.
100 top paths:
0.000454575 unders
0.000341807 dising
0.000293774 pers
0.000285248 cally
0.000232842 inted
0.000204856 heter
0.000198642 tric
0.000188363 ster
0.000185238 hier
0.000183216 unded
0.000174876 heters
0.000169220 sters
0.000153800 stic
0.000142753 pering
0.000132595 dises
0.000131040 ching
0.000129387 shing
0.000125442 dest
0.000115038 teless
0.000113312 resis
0.000112909 dist
0.000112793 dists
0.000112397 mising
0.000109589 sation
0.000109555 fing
0.000108931 dised
0.000107632 ches
0.000102999 cated
0.000101327 dise
0.000097730 sations
0.000096676 inding
0.000095270 nons
0.000093852 lary
0.000093357 ency
0.000091199 ther
0.000090608 ories
0.000090405 alled
0.000088604 apped
0.000088235 hous
0.000087780 trous
0.000082438 undered
0.000081713 noning
0.000079613 dism
0.000079071 thers
0.000078380 ented
0.000076765 irred
0.000076363 disms
0.000074970 geness
0.000074910 veness
0.000071079 vities
0.000070308 mised
0.000068585 ques
0.000068096 unding
0.000067618 sher
0.000067429 peries
\end{verbatim}

\begin{verbatim}
n=5
GetTempFileName failed with error 5
\end{verbatim}

Wikipedia favorites: smally (.000287),
websity (.000156),
stude (.000156),
chool (.000124),
fontry (.000120),
undex (.000102),
octory (.0000990),
coibot (.0000960),
footnot (.0000837),
reporth (.0000518),
delection (.0000484),
grounty (.0000459),
betweek (.0000437),
fination (.0000431),
manuary (.0000388),
whicle (.0000360),
stategory (.0000262),

@@

@@

In the earlier sections I blithely produced tables of the most
probable words according to an $n$-Markov chain. It is not obvious how
to do this (or that it is even possible), so I explain the algorithm
in this section. It's safely skippable, I mean if you don't want to
know about a pretty cool algorithm that's not that complicated and
might even be new, plus {\em dual math}.

Computing the probability of an individual word is easy. We prefix it
with $n$ copies of the start symbol \<, suffix it with a single \>,
and then look up the probability of each symbol given its $n$
preceding symbols in the table, and multiply those all together. We
can compute the probability of any word this way. The problem with
sorting all of the possible words by their probabilities is that there
are an infinite number of them. We can't just look at short words
first, either, because for example the word ``thethethe'' is many
times more likely ($p = 6.08\times 10^{-11}$) than the shorter
``qatzs'' ($9.07\times 10^{-12}$).

The solution is to use coinduction. Most people remember induction
from school, maybe, which is the one where you have some base case
like ``0 is even'', and then you prove that all numbers are either
even or odd by assuming ``$n$ is even or odd'' and proving ``$n + 1$
is even or odd''.

most people remember induction, perhaps 
coinduction: never gonna give you up. once you pop you can't stop.

\section{Backformation}

\begin{verbatim}
Best fixes...
1.000000000 -zzyingly    1/1
1.000000000 -zzying      1/1
1.000000000 -zzuolanas   1/1
1.000000000 -zzuolana    1/1
1.000000000 -zzotints    1/1
1.000000000 -zzotintos   1/1
1.000000000 -zzotinto    1/1
1.000000000 -zzotinting  1/1
1.000000000 -zzotinters  1/1
1.000000000 -zzotinter   1/1
1.000000000 -zzotinted   1/1
1.000000000 -zzotint     1/1
1.000000000 -zzonites    1/1
1.000000000 -zzonite     1/1
1.000000000 -zzoners     1/1
1.000000000 -zzoner      1/1
1.000000000 -zzolans     1/1
1.000000000 -zzolanic    1/1
1.000000000 -zzolan      1/1
1.000000000 -zzmen       1/1
1.000000000 -zzman       1/1
1.000000000 -zzlike      1/1
1.000000000 -zzini       1/1
1.000000000 -zzicatos    1/1
1.000000000 -zzicato     1/1
1.000000000 -zzicati     1/1
1.000000000 -zzian       1/1
\end{verbatim}
1/1 is not really much support.

Ones that are only seen n/1 time never apply (because ...). Filtered:

(actually we should probably be using binomial likelihood)

\begin{verbatim}
Best fixes...
1.000000000 -zzazzes     3/3
1.000000000 -zzazz       3/3
1.000000000 -zzans       3/3
1.000000000 -zzanim      2/2
1.000000000 -zzan        3/3
1.000000000 -zygotic     3/3
1.000000000 -zygotes     2/2
1.000000000 -zygote      2/2
1.000000000 -zygosity    2/2
1.000000000 -zygosis     2/2
1.000000000 -zpas        2/2
1.000000000 -zpahs       2/2
1.000000000 -zpah        2/2
1.000000000 -zpa         2/2
1.000000000 -zooks       2/2
1.000000000 -zonites     3/3
1.000000000 -zonite      3/3
1.000000000 -zonians     2/2
1.000000000 -zonian      2/2
1.000000000 -zists       2/2
1.000000000 -zist        2/2
\end{verbatim}

Not much better!

Only ones with denominator 10+ times.
\begin{verbatim}
Best fixes...
1.000000000 -wrought    10/10
1.000000000 -writings   12/12
1.000000000 -wraps      10/10
1.000000000 -wrap       11/11
1.000000000 -worms      69/69
1.000000000 -worm       69/69
1.000000000 -working    21/21
1.000000000 -worked     14/14
1.000000000 -winds      17/17
1.000000000 -wind       18/18
1.000000000 -wide       12/12
1.000000000 -weights    19/19
1.000000000 -weight     22/22
1.000000000 -waxes      12/12
1.000000000 -wax        12/12
1.000000000 -waters     37/37
1.000000000 -water      41/41
1.000000000 -walking    11/11
1.000000000 -walkers    17/17
1.000000000 -walker     17/17
1.000000000 -valuing    10/10
1.000000000 -valued     11/11
1.000000000 -towns      11/11
1.000000000 -tight      15/15
1.000000000 -thought    12/12
1.000000000 -thinks     11/11
\end{verbatim}

Much better! These are still sort of degenerate and won't help
us find new words because they ALWAYS apply!

Instead anything with p strictly less than 1.
\begin{verbatim}
Best fixes...
0.985714286 -makers     69/70
0.985714286 -maker      69/70
0.983606557 -wood       120/122
0.983471074 -woods      119/121
0.982758621 -down       57/58
0.982658960 -works      170/173
0.981818182 -houses     108/110
0.981818182 -house      108/110
0.981132075 kilo-       52/53
0.980752406 -less       1121/1143
0.980743395 over-       2190/2233
0.980000000 -books      49/50
0.980000000 -book       49/50
0.979591837 -proof      48/49
0.979310345 -lessnesses 142/145
0.979069767 -ships      421/430
0.978723404 -lessness   184/188
0.978723404 -board      138/141
0.978494624 -woman      91/93
0.978021978 -women      89/91
0.977528090 -ship       435/445
0.977272727 -manship    43/44
0.976744186 -weeds      84/86
0.976470588 after-      83/85
0.976190476 -manships   41/42
0.976190476 -making     41/42
0.976190476 -craft      41/42
0.976190476 -boats      41/42
0.976190476 -boat       41/42
\end{verbatim}

Wow, now we're talking!

@@

% TODO: length normalization. just take the average (or max?)
% probability for a word of length n, and normalize by that (which means what?)?
